Deep Reinforcement Learning for Conservation Decisions

### 2021-06-07 manuscript/manuscript.Rmd
Cutting and editing a few sentences

Any application of RL can be divided into two components: an *environment* and an *agent*.
An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
Over the course of a task, an *agent*, which is often a computer program, continuously interacts with the environment. 
At each time step, the agent observes the current *state* of the environment then performs an *action*; as a result of this action, the environment will transition to a new state and will transmit a numerical *reward* signal to the agent.
The goal of the agent is to learn how to maximize its expected cumulative reward.
The agent learns how to achieve this objective during a period called *training*.
In training, the agent *explores* the available states and actions. 
Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.

@@

Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
The training period involves a trial and error process whereby the agent will perform an action and then receive feedback from the environment in the form of a reward. 
This process will repeat until the environment transitions to a terminal state, after which the environment will reset to an initial state.
According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its policy.
The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
An advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing anything about the environment's transition dynamics or the inner workings of the algorithm.

@@

In this brief section, we will introduce the concept of the RL environment as partially observable Markov decision process (POMDP); we will then use POMDP definitions to define the RL objective.
Since all RL algorithms are either directly or indirectly maximizing the RL objective, understanding the RL objective is critical to understanding how RL algorithms work. <!-- ML: this seems like an obvious sentence to me but am stuck on it atm so revisions welcomed-->

@@

The agent interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
At each time step, the agent makes an observation of the environment at time t, $o_t$, and selects an action, $a_t$, according to its policy, $\pi$, which maps observations to a probability distribution over actions, $\pi: \Omega \rightarrow p(\mathcal{A} \mid \Omega)$.
The action will cause the environment to transition to a new state according to the state transition operator. 
The agent will then receive a new observation and a discounted reward signal, $\gamma^t \, r(s_t, a_t)$, as feedback. 
This process will repeat until a terminal state is reached.
As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
From these definitions, we can provide an agent's trajectory distribution for a given policy as,

@@

Note that this formalism extends to an infinite horizon case when $T= \infty$.
While there are RL methods to deal with infinite horizon cases, we will only present methods for finite horizon or episodic POMDPs.
In the appendix, we will go into further detail on the general approaches that model-free deep RL algorithms take towards optimizing the RL objective.

@@

To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.

Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
Generally, model-free agents are either learning a policy function, a value function or both.
For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
The class of algorithms that exclusively try to learn a value function are called *value-based* methods.
In contrast, *policy gradient* algorithms exclusively learn the policy.
And, lastly, *actor-critic* algorithms attempt to learn both a policy and value function, whereby the value function is used to inform the agent on the goodness of a selected action.
We go into greater depth on these different classes of RL algorithms in Appendix A.

Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
Classic RL algorithms, like policy iteration, value iteration and TD-learning, fail when the state-action space is large, e.g. when the possible states and actions can not be represented in a tractable table [@suttonbarto].
Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments [@DQN, @alphaGo2016].
Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@universalapproximators].
Whenever there is a need for an agent to approximate some function, say a policy function or value function, neural networks can be used in this capacity.
While there are other function approximators that can be used in RL, e.g. Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network, the neural network that the agent is using to approximate the policy function, so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future. 
This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively

@@

On top of the aforementioned distinctions for model-free, deep RL algorithms, there are more algorithm design choices that can be made. 
One important point of divergence is whether the agent learns a deterministic or stochastic policy.
Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
In table 1, we provide some of the commonly used RL algorithms and delineate some of their important characteristics. <!-- ML: Have we dropped the table? -->
In Appendices A and B, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.

@@

## Sustainable harvest \label{sec:fishery}

@@

## Ecological Tipping Points \label{sec:tipping}

@@

## Additional Environments

Ecology holds many open problems for deep RL.  
It is straight-forward to extend the simple environments presented here to reflect greater biological complexity, or to consider a host of more realistic decision scenarios.
We provide an initial library of example environments at <https://boettiger-lab.github.io/conservation-gym>.
Some environments in this library include a wildfire `gym` that poses the problem on wildfire suppression with a cellular automata model, an epidemic `gym` that examines timing of interventions to curb disease spread, as well as more complex variations of the fishing and conservation environments presented above.

>>>>

Any application of RL can be divided into two components: an *environment* and an *agent*.
An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
The *agent*, which is often a computer program, continuously interacts with the environment. 
At each time step, the agent observes the current *state* of the environment then performs an *action*; as a result of this action, the environment will transition to a new state and will transmit a numerical *reward* signal to the agent.
The goal of the agent is to learn how to maximize its expected cumulative reward.
The agent learns how to achieve this objective during a period called *training*.
In training, the agent *explores* the available states and actions. 
Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.

@@

Training a deep RL agent on an environment involves allowing the agent to interact with the environment for thousands or even millions of time steps.
The training period involves a trial and error process in which the agent inputs actions to the environment and uses the reward signal to learn an optimal policy. 
According to the algorithm that the agent follows, the agent will explore the state-action space and use its experience to improve its decisions.
The amount of time needed for an agent to learn high reward yielding behavior cannot be predetermined and depends on a host of factors including the amount of dedicated compute, complexity of the environment, complexity of the agent and more.
Yet, overall, it has been well established that deep, model-free RL agents tend to be very sample inefficient [@gu_q-prop_2017], so it is recommended to provide a generous training budget for these agents.
An advantage of using deep, model-free algorithms is that since they function autonomously and do not require any knowledge about the environment, we can readily test how these algorithms perform without knowing anything about the environment's transition dynamics or the inner workings of the algorithm.

@@

In this brief section, we will introduce the concept of the RL environment as a partially observable Markov decision process (POMDP); we will then use POMDP definitions to define the RL objective.

@@

The agent interacts with the environment in an iterative loop, whereby the agent only has access to the observation space, action space and the discounted reward signal, $\gamma^t \, r(s_t, a_t)$.
At each time step, the agent makes an observation of the environment at time t, $o_t$, and selects an action, $a_t$, according to its policy, $\pi$, which maps observations to a probability distribution over actions, $\pi: \Omega \rightarrow p(\mathcal{A} \mid \Omega)$.
The action will cause the environment to transition to a new state according to the state transition operator. 
The agent will then receive a new observation and a discounted reward signal, $\gamma^t \, r(s_t, a_t)$, as feedback. 
This process will repeat until a terminal state is reached, after which the environment can reset to an initial state.
As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the length of the state-action sequence.
From these definitions, we can provide an agent's trajectory distribution for a given policy as,

@@

Note that this formalism extends to an infinite horizon case when $H= \infty$.
While there are RL methods to deal with infinite horizon cases, we will only present methods for finite horizon or episodic POMDPs.
In the appendix, we will go into further detail on the general approaches that model-free deep RL algorithms take towards optimizing the RL objective.

@@

To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment; yet, *model-based* algorithms employ a model of the environment to achieve the RL objective.
A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@mbpo].
Since our examples use model-free agents, we will focus on model-free algorithms in this section.

Among model-free methods, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
Generally, model-free agents are either learning a policy function, a value function or both.
For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
The class of algorithms that exclusively try to learn a value function are called *value-based* methods.
In contrast, *policy gradient* algorithms exclusively learn the policy.
And, lastly, *actor-critic* algorithms attempt to learn both a policy and value function, whereby the value function is used to inform the agent on the goodness of a selected action.
We go into greater depth on these different classes of RL algorithms in Appendix A.

Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
Classic RL algorithms, like policy iteration, value iteration and TD-learning, fail when the state-action space is large, e.g. when the possible states and actions can not be represented in a tractable table [@suttonbarto].
Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments [@DQN, @alphaGo2016].
Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@universalapproximators].
Whenever there is a need for an agent to approximate some function, say a policy function or value function, neural networks can be used in this capacity.
While there are other function approximators that can be used in RL, e.g. Gaussian processes [@grande14], neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@briefsurveydrl].
We will not discuss in detail how neural networks work -- see [@lecun_deep_2015] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy-approximating neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future. 
This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively

@@

On top of the aforementioned distinctions for model-free, deep RL algorithms, there are more algorithm design choices that can be made. 
One important point of divergence is whether the agent learns a deterministic or stochastic policy.
Another is whether the algorithm is formulated to work on a discrete, continuous or both a discrete and continuous state and action space.
In Appendices A and B, we will introduce the considerations one should make in deciding whether an algorithm is well suited for a specific environment.

@@

## Sustainable harvest

@@

## Ecological Tipping Points

@@

## Unsolved environments: Wildfires and other examples

<!-- ML: Need to rework this to talk about other environments that we are working on basically point to conservation gym website -->
One feature that most distinguishes RL from other approaches to decision theory (such as optimal control) is it's flexibility or ambition.
While optimal control approaches typically require extensive simplifying assumptions such as discrete states or actions, deterministic dynamics or short time horizons, RL is scoped more ambitiously.
Large, complex problems are still immensely challenging to RL, and existing RL approaches will frequently struggle to solve problems that can already be solved much more efficiently by other means.
Nevertheless, deep RL approaches using neural network architectures have proven successful on problems that no amount of computational power would put into reach of purely conventional methods.
Our previous examples have focused on relatively simple examples that can also be addressed at least approximately by conventional methods, and also permit RL-based solutions to be found relatively efficiently on commodity hardware.
To better illustrate the generality of RL approach and more typical complexity of conservation problems, we now introduce a third example for which we have not yet found an effective policy using RL or classical methods.

<!-- Marcus add ~ paragraph discussion of the fire gym, citing motivating examples, basic overview of the state space, action space and reward and a screenshot of the simulation as a figure? -->
